Live Functional Programming with Typed Holes

### 2018-04-16 intro.tex
unbury the phrase 'this paper develops a dynamic semantics for incomplete functional programs'; rephrase more occurences of 'is'

For example, read-evaluate-print loops (REPLs) and derivatives thereof, like the IPython/Jupyter lab notebooks popular in data science~\cite{PER-GRA:2007}, allow the programmer to edit and immediately execute program fragments organized into a sequence of cells. 
Spreadsheets are live functional dataflow environments, with cells organized into a grid \cite{DBLP:journals/jfp/Wakeling07}. 
More specialized examples include live direct manipulation programming environments like SuperGlue
\cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist}, and the tools
demonstrated by \citet{victor2012inventing} in his lectures;
%
live user interface frameworks \cite{burckhardt2013s};
%
live image processing languages~\cite{DBLP:journals/vlc/Tanimoto90};
%
and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding in an artistic performance.
Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing the state of a running program, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments. 
Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13}, has been and continues to be an active area of research and development.

%The problem that specifically motivates this paper \IS that
Meanwhile,
programming languages typically assign dynamic meaning only to {complete programs}, i.e. programs that are syntactically well-formed and free of static type and binding errors. A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states. As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e. it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer \IS actually seeing and writing in the editor.\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}

In some cases these gaps are momentary, like while the programmer
%\IS entering
enters
a short expression. In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, or when the programmer makes a mistake (so they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}).
%
The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program. Throughout the process of addressing these errors, the program text remains formally meaningless. 
Overall, an analysis of edits performed by Java programmers using Eclipse found that about 40\% of edit states were malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.

@@

In recognition of this problem---that incomplete programs lack meaning---%
\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
functional programs, modeled formally as typed expressions with \emph{holes} in 
both expression and type position. 
Empty holes stand for missing expressions or types,
and non-empty holes operate as ``membranes'' around static type inconsistencies 
(i.e. they internalize the ``red underline'' that editors commonly display under a type inconsistency).
% For editor states into this language of incomplete programs, 
% editor services can reason about types and binding in many more situations than previously possible.
\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from an editor state. 
Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
the programmer can insert explicitly holes manually (see below) 
or semi-automatically via code completion techniques \cite{Amorim2016}, 
or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).

For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes.
%
%Our goal in this paper \IS to develop
This paper develops
just such a dynamic semantics for incomplete functional programs, 
based on the static semantics just mentioned.

One approach would be for evaluation to simply stop with an error when it encounters an expression hole.
This mirrors a workaround that programmers commonly deploy: 
raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
Some languages provide more concise notation for this interpretation of expression holes, 
e.g. Scala uses \lismall{???} for missing method bodies 
and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u} where \lismall{u} \IS an optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
\footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}

@@

Moreover, we need to be able to run programs containing not only expression holes but also type holes.
In the static semantics developed by \citet{popl-paper}, the machinery for 
reasoning statically about type holes \IS derived from gradual type theory, 
identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}. 
To specify our intended dynamic semantics for incomplete programs, then, we can look to the dynamic semantics from
gradual type theory, which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
an exception. As a result, the live programming environment \IS left unable to provide feedback about even parts 
of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).

>>>>

For example, read-evaluate-print loops (REPLs) and derivatives thereof, like the IPython/Jupyter lab notebooks popular in data science~\cite{PER-GRA:2007}, allow the programmer to edit and immediately execute program fragments organized into a sequence of cells. 
Spreadsheets are live functional dataflow environments, with cells organized into a grid \cite{DBLP:journals/jfp/Wakeling07}. 
More specialized examples include live direct manipulation programming environments like SuperGlue
\cite{McDirmid:2007}, \sns{}~\cite{sns-pldi,sns-uist}, and the tools
demonstrated by \citet{victor2012inventing} in his lectures;
%
live user interface frameworks \cite{burckhardt2013s};
%
live image processing languages~\cite{DBLP:journals/vlc/Tanimoto90};
%
and live visual and auditory dataflow languages \cite{DBLP:conf/vl/BurnettAW98}, which can support live coding in an artistic performance.
Editor-integrated debuggers \cite{mccauley2008debugging} and other systems that support editing the state of a running program, like Smalltalk environments \cite{Goldberg:1983cn}, are also live programming environments. 
Live programming, in its various incarnations \cite{DBLP:journals/vlc/Tanimoto90,DBLP:conf/icse/Tanimoto13},
%has been and continues to be an active area of research and development.
has been, and continues to be, an active area of research and development.

% \matt{ ``The problem that specifically motivates this paper'' is a
% really long noun phrase; the sentence is more interesting, IMO, when
% the subject is 'programming languages' and not that noun phrase.  }
%
%The problem that specifically motivates this paper \IS that
Problematically,
programming languages typically assign dynamic meaning only to {complete programs}, i.e. programs that are syntactically well-formed and free of static type and binding errors. A program editor, however, frequently encounters incomplete, and therefore meaningless, editor states. As a result, live feedback either ``flickers in and out'', creating temporal gaps, or it ``goes stale'', i.e. it relies on the most recent complete editor state, creating a perceptive gap because the feedback does not necessarily reflect what the programmer
%
%\IS actually seeing and writing in the editor.
actually sees and writes in the editor.
\todo{Could put the sentence about ``this \IS when live feedback would be most useful'' here, but thats an empirical statement that we do not support...}

In some cases these gaps are momentary, like while the programmer
%\IS entering
enters
a short expression. In other cases, these gaps can persist over substantial lengths of time, such as when there are many branches of a case analysis whose bodies are initially left blank, when the programmer changes a type definition, or when the programmer makes a mistake;
Consequently, they arise somewhat more frequently for novice programmers \cite{mccauley2008debugging,fitzgerald2008debugging}.
%
%\matt{we are not really modeling a language with user type definitions -- what does the following sentence really add here? also, I just added this scenerio to the list above}
%The problem \IS particularly pronounced in languages with rich static type systems, where certain program changes, such as a change to a type definition, can cause type errors to propagate throughout the program. Throughout the process of addressing these errors, the program text remains formally meaningless. 
Overall, an analysis of edits performed by Java programmers using Eclipse found that about 40\% of edit states were malformed \cite{popl-paper,6883030} and some additional number, which could not be determined from the data, were well-formed but ill-typed.

@@

%\matt{ Below, notice the broadening from ``the problem that motivates
%this paper is that incomplete lack dynamic meaing'' to just the
%general quesiton of ``incomplete program meaning''--- another reason
%to do my edit above, and remove that specific phrase.
%}
In recognition of this problem---that incomplete programs lack meaning---%
\citet{popl-paper} developed a static semantics (i.e. a type system) for incomplete 
functional programs, modeled formally as typed expressions with \emph{holes} in 
both expression and type position. 
Empty holes stand for missing expressions or types,
and non-empty holes operate as ``membranes'' around static type inconsistencies 
(i.e. they internalize the ``red underline'' that editors commonly display under a type inconsistency).
% For editor states into this language of incomplete programs, 
% editor services can reason about types and binding in many more situations than previously possible.
\citet{popl-paper,HazelnutSNAPL} discuss several ways to determine an incomplete expression from an editor state. 
Briefly, error recovery mechanisms can insert the necessary holes implicitly \cite{DBLP:journals/siamcomp/AhoP72,charles1991practical,graham1979practical,DBLP:conf/oopsla/KatsJNV09,DBLP:conf/oopsla/KatsV10},
the programmer can insert explicitly holes manually (see below) 
or semi-automatically via code completion techniques \cite{Amorim2016}, 
or a structure editor can insert holes fully automatically \cite{popl-paper} (we say more about this technique in Sec.~\ref{sec:implementation}).
%
For the purposes of live programming, however, a static semantics for incomplete programs does not suffice---%
we also need a corresponding dynamic semantics for incomplete programs that specifies how to evaluate expressions with holes.
%
%Our goal in this paper \IS to develop

%%%%%%%%%%%% New par:
To address this need,
this paper develops
a \emph{dynamic semantics} for incomplete functional programs.
%
To understand this problem, consider a (common) approach in practice, where evaluation simply stops with an error when it encounters an ``expression hole,'' encoded by the programmer with a raised exception.
%This mirrors a workaround that programmers commonly deploy: 
%raising an exception when the intended expression has yet be determined, e.g. \lismall{raise Unimplemented}. 
Some languages provide more concise notation for this interpretation of expression holes, 
e.g. Scala uses \lismall{???} for missing method bodies 
and GHC Haskell with the \lismall{-fdefer-typed-holes} or \lismall{-fdefer-ty}\lismall{pe-errors} flags uses \lismall{_u}
%where \lismall{u} \IS an
where \lismall{u} serves as an
optional hole name \cite{GHCWIKI}\todo{fix citation}{}.%
\footnote{Without these flags, holes cause compilation to fail. The compiler reports static information about each hole's type and typing context. 
%This \IS also how holes in proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave.}
proof assistants like Agda \cite{norell:thesis,norell2009dependently} and Idris \cite{brady2013idris} behave this way.}

@@

Moreover, we need to be able to run programs containing not only expression holes but also type holes.
In the static semantics developed by \citet{popl-paper}, we derive the machinery for 
reasoning statically about type holes from gradual type theory, 
identifying the type hole with the unknown type \cite{DBLP:conf/snapl/SiekVCB15,Siek06a}. 
To specify our intended dynamic semantics for incomplete programs, then, we can look to the dynamic semantics from
gradual type theory, which selectively inserts dynamic casts as necessitated by missing type information to maintain type safety. 
%However, this \IS again somewhat dissatisfying from the perspective of live programming because when a cast fails, evaluation again stops with 
However, when a cast fails, evaluation again stops with 
an exception.
%
As a result, traditional gradual typing approaches still leave the live programming environment unable to provide feedback about the parts 
of the program that do not depend on the problematic sub-expression (e.g. later cells in a lab notebook).
